---
title: "Palmer Penguins Visualisation"
author: "ChatGPT"
output:
  bookdown::html_document2
---

## Introduction

The `palmerpenguins` dataset contains measurements of penguins from
different species and islands. The following visualisations explore
relationships and differences between penguin species.

## 1. Scatter Plot


Figure @fig-scatter-plot shows the relationship between bill length
and bill depth. The different species form distinct groups.

```{r fig-scatter-plot, echo=TRUE, fig.cap="Bill length and bill depth by penguin species."}
library(ggplot2)
library(palmerpenguins)

ggplot(penguins, aes(x = bill_length_mm, 
                     y = bill_depth_mm, 
                     colour = species)) +
  geom_point() +
  labs(
    x = "Bill Length (mm)",
    y = "Bill Depth (mm)",
    colour = "Species"
  ) +
  theme_minimal()
```



## 2. Box Plot

Figure @fig-box-plot compares the distribution of body mass across
the three penguin species. Gentoo penguins generally have a higher body
mass than the other species.

```{r fig-box-plot}
ggplot(penguins, aes(x = species, y = body_mass_g)) +
  geom_boxplot() +
  labs(
    x = "Species",
    y = "Body Mass (g)"
  ) +
  theme_minimal()
```

## 3. Histogram

Figure @fig-histogram shows the distribution of penguin body mass.
The distribution contains substantial variation, reflecting differences
between species.

```{r fig-histogram}
ggplot(penguins, aes(x = body_mass_g)) +
  geom_histogram(bins = 15) +
  labs(
    x = "Body Mass (g)",
    y = "Count"
  ) +
  theme_minimal()
```

## 4. Bar Chart

Figure @fig-bar-chart shows the number of observations for each
penguin species. Adelie has the largest number of observations in the
dataset.

```{r fig-bar-chart} 
ggplot(penguins, aes(x = species)) +
  geom_bar() +
  labs(
    x = "Species",
    y = "Number of Penguins"
  ) +
  theme_minimal()
```

## 5. Faceted Scatter Plot

Figure @fig-facetedPlot shows the relationship between flipper
length and body mass separately for each penguin species. A positive
relationship is visible within each species.

```{r fig-facetedPlot}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point() +
  facet_wrap(~species) +
  labs(
    x = "Flipper Length (mm)",
    y = "Body Mass (g)"
  ) +
  theme_minimal()
```